---
output: 
  github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.path = "results/Readme_figures/",
  message = FALSE,
  dpi = 300
)
set.seed(1)
```

# TinGa

This is a repository hosting the R implementation of TinGa [1], a trajectory inference method based on the Growing Neural Gas algorithm [2].


## Getting Started

These instructions show how to use TinGa on an example single cell dataset and how to visualise its results using the dynplot package hosted on github (see https://github.com/dynverse/dynplot)

### Installing

TinGa can be used in the dynverse framework (see https://dynverse.org/users/2-quick_start/), which offers a wide range of functions for visualisation of trajectories and genes of interest in single cell datasets. 

TInGa, dynwrap and dynplot can be installed in R as follows:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("Helena-todd/TInGa/package")
```

### Usage of TInGa, together with dynwrap and dynplot

Any single cell dataset can be wrapped into an object for further use with dyn- packages. Two matrices need to be provided: 
* the counts matrix, of the form cells * features
* the normalised expression matrix, of the form cells * features

We provide an example dataset 

```{r}
library(TInGa)
library(tidyverse)
library(dyno)

data <- wrap_expression(
  expression = TInGa::data_exp,
  counts = TInGa::data_counts
)
```

Apply TinGa to the data

```{r}
traj <- infer_trajectory(
  data, 
  method = gng_param2(), 
  seed = 42
)
```

Visualise the resulting trajectory using dynplot

```{r}
plot_dimred(
  traj,
  label_milestones = TRUE
)
```

The features that vary the most along the trajectory can be identified and plotted in the form of a heatmap using dynplot. If the trajectory contains branching, the different branches will be represented separately in the heatmap

```{r}
plot_heatmap(
  traj, 
  expression_source = data$expression
)
```

Visualise the expression of one feature of interest in the trajectory

```{r}
plot_dimred(
  traj, 
  color_cells = "feature", 
  expression_source = data$expression, 
  feature_oi = "MT1E"
)
```

## TinGa with a prior dimensionality reduction
Use the following code to run TinGa on a custom dimensionality reduction (e.g. UMAP, tSNE, PCA, ...) provided by you.

```{r}
# calculate a PCA
dimred <- prcomp(TInGa::data_exp)$x[,1:2]

data_with_prior <-
  wrap_expression(
    expression = TInGa::data_exp,
    counts = TInGa::data_counts
  ) %>%
  add_prior_information(
    dimred = dimred
  )
```

Apply TinGa to custom dimred

```{r}
traj <- infer_trajectory(
  data_with_prior, 
  method = gng_param2(), 
  give_priors = "dimred",
  seed = 42
)
```

Visualise the resulting trajectory using dynplot

```{r}
plot_dimred(
  traj,
  label_milestones = TRUE
)
```

## Acknowledgments

* to @zouter and @rcannood for the remarkable dynverse packages they created
* to @rcannood for an initial implemenation of the GNG algorithm

## References

[1] Helena Todorov, Robrecht Cannoodt, Wouter Saelens, Yvan Saeys (2020) TinGa: fast and flexible trajectory inference with Growing Neural Gas, _Bioinformatics_ 36 i66–i74, https://doi.org/10.1093/bioinformatics/btaa463.

[2]: Fritzke, B. (1995) A growing neural gas network learns topologies, _Advances in Neural Information Processing Systems_ 7, 625-632.
